Supporting Information for "Imperfect slope measurements drive overestimation in geometric cone model of lake and reservoir depth"

### 2021-11-10 manuscript/manuscript.Rmd
stash progress

One intuitive approach for predicting waterbody depth involves using a geometric model that assumes waterbody basins correspond to an idealized shape such as a cone, bowl, or an elliptic sinusoid \citep{getiranaDerivingThreeDimensional2018, hollisterPredictingMaximumLake2011, neumannMaximumDepthAverage1959, yigzawNewGlobalStorage2018}. All such geometric models for waterbody depth prediction involve implicit assumptions about the terms of geometric formulae. In the simplest case, where waterbody basins are treated as cones (Eq. \ref{eq1}, Fig. 1), two assumptions are required to make depth predictions for all waterbodies: 1) that nearshore land slope is a representative proxy for in-lake slope and 2) that the distance to the center of the waterbody is a representative proxy for the distance to the deepest point of the waterbody (Fig. 1). This cone model imposes the following fixed (i.e. geometric) relationship between slope and horizontal distance:

@@

where the product of slope and horizontal distance yields an exact geometric depth estimate ($depth_{geometric}$). Cone models of waterbody basins have been used extensively to estimate hypsography in waterbodies where there is no knowledge of volume or mean depth (Read et al. 2014, Winslow et al. 2017).

@@

By definition, the distance proxy (distance to the center of the waterbody) must always be greater or equal to the true distance value (distance to the deepest point of the waterbody). Therefore, we expect that the use of this proxy will lead to overestimation of waterbody depth (Fig. 1). Furthermore, we expect to see greater overestimation error in reservoirs as compared to natural lakes because many reservoirs are known to be drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir (Lanza and Silvey 1985). In a similar fashion, we expect to see overestimation error associated with using a nearshore land slope proxy in waterbodies with differing cross-section shape such that the depth of bowl-shaped (i.e. concave) waterbodies will be overpredicted whereas the depth of V-shaped (i.e. convex) waterbodies will be underpredicted (Fig. S1). Finally, we expect that depth predictions themselves will be strongly related to waterbody area and hydrologic subbasin variables as these measures have been influential in prior studies \citep{oliver2016prediction}.

@@

We used our generated bathymetry surfaces to find the location of the deepest point in the waterbody and we resolved ties by choosing the deepest point that was closest to the center of the waterbody. We used the location of this deepest point to calculate "distance to the deepest point" as the minimum distance to the waterbody shoreline. To account for waterbodies where the centroid does not intersect waterbody bathymetry because it is located within an embedded island or peninsula, we calculated the center of the waterbody not as its centroid but rather by finding the point farthest from the waterbody shoreline (i.e. its “visual distance to waterbody center”). For these calculations, we used the polylabelr R package (Larsson 2019), which interfaces with the Mapbox pole of inaccessibility algorithm (Agafonkin 2019). We calculated (maximum) in-lake slope as depth at the deepest point divided by the shortest distance to the deepest point from the shoreline.. We calculated (mean) nearshore land slope for each waterbody by computing the slope within a 100-m buffer using data from a high-resolution digital elevation model (~ 15x15m grain) accessed using the elevatr R package (Hollister and Shah 2017). We explored alternative buffer sizes ranging between 50-1000 m following Sobek et al. (2011) and although 100 m provided the lowest model error, we ultimately found little appreciable effect of varying buffers on model performance. Slope computations proceeded by passing a 3x3 moving window over the 100-m buffer to calculate the slope at each point using Horn's algorithm via the terrain function in the raster R package (Hijmans 2019). Reported nearshore land slope values are the mean of all points in the buffer. In addition to the aforementioned techniques of calculating in-lake (and nearshore) slopes and distances, we tried 7 alternate techniques which are described in Fig. S5 and Table S2 including measures such as median slope (results not shown).

@@

## Model description
### Geometric model 
We used a geometric model of waterbodies where basins are treated as cones with a fixed relationship between slope and distance (Equation \ref{eq1}). One reason that we used the cone model is that, unlike other idealized shapes, it does not require any knowledge of waterbody volume or mean depth. Note that Equation \ref{eq1} is a geometric formula and has no intercept or coefficients and it produces an exact depth value given true values of slope and distance. To use this model to predict the depth of all waterbodies, there is a necessary assumption that proxy slope and distance measures, which are available for all waterbodies, are representative of true slope and distance (Fig. 1).

@@

## Results
Waterbodies belonging to each cross-section shape and reservoir class were not evenly distributed across our study area (Fig. 2). For example, concave waterbodies were nearly absent from Michigan whereas Maine had an overabundance (~3%) of waterbodies categorized as neither concave nor convex. Waterbodies in the southern portions of our study area tended to be classified as reservoirs whereas waterbodies in the northern portions of our study area were a more even mix between reservoirs and natural lakes (Fig. 2). Approximately 18%, 80%, and 2% of waterbodies were classified as having a concave, convex, or neither shape respectively whereas approximately 30% and 70% of waterbodies were classified as being a reservoir or a natural lake.

Although proxy distance to waterbody center was often larger in magnitude compared to the true distance to the deepest point of waterbodies’ (rather than being identical), they were strongly related ($R^2$ = 0.8). Note that the coefficient of determination for this relationship is not strictly correct given that distance to waterbody center is an upper bound on distance to the deepest point of waterbodies. In contrast to distance metrics, proxy nearshore land slope and true in-lake slope were more weakly related ($R^2$ = 0.17). For slope measures, most waterbodies had higher magnitude (i.e. steeper) nearshore land slope compared to true in-lake slope (Fig. 3). Taken together, these results suggest that proxy distance to the center of waterbodies is representative of true distance to the deepest point of waterbodies whereas proxy nearshore land slope is not representative of true in-lake slope. The strong relationship between distance to the center of waterbodies and distance to the deepest point means that it is possible to convert between the two measures in subsequent analyses (See best-fit equations in Fig. 3).

@@

Furthermore, analysis of model residuals showed overestimation of waterbody depth for concave waterbodies when models included a proxy slope measure (Fig. 4). We observed similar but smaller overestimation depending on if a waterbody was classified as a reservoir rather than a natural lake (Fig. 4). We found that models restricted to consider only concave lakes had lower error (both in absolute and relative terms) compared to models on other data subsets (e.g. convex lakes, reservoirs, natural lakes, see Table 2). Conversely, we did not observe any notable geographic patterns in model residuals (Fig. S6).

@@

The most important covariates in these models were those relating to spatial location, waterbody area, and perimeter (Fig. 5). Conversely, watershed metrics and waterbody elevation had little contribution to random forest model fit. The spatial location (i.e. HUC4, hydrologic subbasin) covariate was notably less important in the true slope model compared to the two proxy slope models. To evaluate the contribution of our “offset” models relative to the “base” geometric model, we can look at model importance calculations for the geometric max depth input to the random forest model (Fig. 5). These calculations indicate that omitting a geometric max depth term results in a 130%, 60%, or 50% increase in mean square error depending on the formulation of geometric max depth in Eq. \ref{eq1}.

@@

We found other differences among waterbody geometry measures according to waterbody cross-section shape. One finding was that convex waterbodies, when compared to concave waterbodies, had longer distances to waterbody centers relative to corresponding distances to the deepest point of waterbodies. In addition, convex waterbodies often had steeper in-lake slopes relative to nearshore land slopes as compared to concave waterbodies. Finally, it was notable that convex waterbodies were deeper than concave waterbodies despite having similar distributions of waterbody surface area (Fig. S8). The underlying cause of these differences is unknown but one possibility is that geometry is tied to the circumstances of waterbody formation whereby the formation of concave waterbodies were a result of more intense glacial scouring compared to that of convex waterbodies \citep{gorhamPhysicalLimnologyNorthern1958}. While our findings provide some evidence in support of this idea, namely that there is a geographic hotspot of concave waterbodies associated with the glaciated “prairie pothole region” (see \citealt{hayashiSimpleEquationsRepresent2000}), the overall geographic distribution of waterbody cross-section shapes does not support this idea. Instead of a concentrated area of concave waterbodies in formerly glaciated regions, there appears to be a fairly even mix of concave and convex waterbodies distributed amongst the northern (i.e. glaciated) and southern (non-glaciated) portions of our study area (Fig. 2).

>>>>

One intuitive approach for predicting waterbody depth involves using a geometric model that assumes waterbody basins correspond to an idealized shape such as a cone, bowl, or an elliptic sinusoid \citep{getiranaDerivingThreeDimensional2018, hollisterPredictingMaximumLake2011, neumannMaximumDepthAverage1959, yigzawNewGlobalStorage2018}. All such geometric models for waterbody depth prediction involve implicit assumptions about the terms of geometric formulae. In the simplest case, where waterbody basins are treated as cones (Eq. \ref{eq1}, Fig. \ref{fig1}), two assumptions are required to make depth predictions for all waterbodies: 1) that nearshore land slope is a representative proxy for in-lake slope and 2) that the distance to the center of the waterbody is a representative proxy for the distance to the deepest point of the waterbody (Fig. \ref{fig1}). This cone model imposes the following fixed (i.e. geometric) relationship between slope and horizontal distance:

@@

where the product of slope and horizontal distance yields an exact geometric depth estimate ($depth_{geometric}$). Cone models of waterbody basins have been used extensively to estimate hypsography in waterbodies where there is no knowledge of volume or mean depth \citep{readSimulating2368Temperate2014, winslowLargescaleModeledContemporary2017}.

@@

By definition, the distance proxy (distance to the center of the waterbody) must always be greater or equal to the true distance value (distance to the deepest point of the waterbody). Therefore, we expect that the use of this proxy will lead to overestimation of waterbody depth (Fig. \ref{fig1}). Furthermore, we expect to see greater overestimation error in reservoirs as compared to natural lakes because many reservoirs are known to be drowned river valleys where the deepest point is close to the edge at the end of the reservoir (i.e. next to the dam) rather than in the center of the reservoir \citep{lanza1985interactions}. In a similar fashion, we expect to see overestimation error associated with using a nearshore land slope proxy in waterbodies with differing cross-section shape such that the depth of bowl-shaped (i.e. concave) waterbodies will be overpredicted whereas the depth of V-shaped (i.e. convex) waterbodies will be underpredicted (Fig. S1). Finally, we expect that depth predictions themselves will be strongly related to waterbody area and hydrologic subbasin variables as these measures have been influential in prior studies \citep{oliver2016prediction}.

@@

We used our generated bathymetry surfaces to find the location of the deepest point in the waterbody and we resolved ties by choosing the deepest point that was closest to the center of the waterbody. We used the location of this deepest point to calculate "distance to the deepest point" as the minimum distance to the waterbody shoreline. To account for waterbodies where the centroid does not intersect waterbody bathymetry because it is located within an embedded island or peninsula, we calculated the center of the waterbody not as its centroid but rather by finding the point farthest from the waterbody shoreline (i.e. its “visual distance to waterbody center”). For these calculations, we used the polylabelr R package \citep{larssonPolylabelrFindPole2019}, which interfaces with the Mapbox pole of inaccessibility algorithm \citep{agafonkinJSLibraryFinding2019}. We calculated (maximum) in-lake slope as depth at the deepest point divided by the shortest distance to the deepest point from the shoreline.. We calculated (mean) nearshore land slope for each waterbody by computing the slope within a 100-m buffer using data from a high-resolution digital elevation model (~ 15x15m grain) accessed using the elevatr R package \citep{hollisterElevatrAccessElevation2017}. We explored alternative buffer sizes ranging between 50-1000 m following \citet{sobekPredictingDepthVolume2011} and although 100 m provided the lowest model error, we ultimately found little appreciable effect of varying buffers on model performance. Slope computations proceeded by passing a 3x3 moving window over the 100-m buffer to calculate the slope at each point using Horn's algorithm via the terrain function in the raster R package \citep{hijmansRasterGeographicData2019}. Reported nearshore land slope values are the mean of all points in the buffer. In addition to the aforementioned techniques of calculating in-lake (and nearshore) slopes and distances, we tried 7 alternate techniques which are described in Fig. S5 and Table S2 including measures such as median slope (results not shown).

@@

## Model description
### Geometric model 
We used a geometric model of waterbodies where basins are treated as cones with a fixed relationship between slope and distance (Equation \ref{eq1}). One reason that we used the cone model is that, unlike other idealized shapes, it does not require any knowledge of waterbody volume or mean depth. Note that Equation \ref{eq1} is a geometric formula and has no intercept or coefficients and it produces an exact depth value given true values of slope and distance. To use this model to predict the depth of all waterbodies, there is a necessary assumption that proxy slope and distance measures, which are available for all waterbodies, are representative of true slope and distance (Fig. \ref{fig1}).

@@

## Results
Waterbodies belonging to each cross-section shape and reservoir class were not evenly distributed across our study area (Fig. \ref{fig2}). For example, concave waterbodies were nearly absent from Michigan whereas Maine had an overabundance (~3%) of waterbodies categorized as neither concave nor convex. Waterbodies in the southern portions of our study area tended to be classified as reservoirs whereas waterbodies in the northern portions of our study area were a more even mix between reservoirs and natural lakes (Fig. \ref{fig2}). Approximately 18%, 80%, and 2% of waterbodies were classified as having a concave, convex, or neither shape respectively whereas approximately 30% and 70% of waterbodies were classified as being a reservoir or a natural lake.

Although proxy distance to waterbody center was often larger in magnitude compared to the true distance to the deepest point of waterbodies’ (rather than being identical), they were strongly related ($R^2$ = 0.8). Note that the coefficient of determination for this relationship is not strictly correct given that distance to waterbody center is an upper bound on distance to the deepest point of waterbodies. In contrast to distance metrics, proxy nearshore land slope and true in-lake slope were more weakly related ($R^2$ = 0.17). For slope measures, most waterbodies had higher magnitude (i.e. steeper) nearshore land slope compared to true in-lake slope (Fig. \ref{fig3}). Taken together, these results suggest that proxy distance to the center of waterbodies is representative of true distance to the deepest point of waterbodies whereas proxy nearshore land slope is not representative of true in-lake slope. The strong relationship between distance to the center of waterbodies and distance to the deepest point means that it is possible to convert between the two measures in subsequent analyses (See best-fit equations in Fig. \ref{fig3}).

@@

Furthermore, analysis of model residuals showed overestimation of waterbody depth for concave waterbodies when models included a proxy slope measure (Fig. \ref{fig4}). We observed similar but smaller overestimation depending on if a waterbody was classified as a reservoir rather than a natural lake (Fig. \ref{fig4}). We found that models restricted to consider only concave lakes had lower error (both in absolute and relative terms) compared to models on other data subsets (e.g. convex lakes, reservoirs, natural lakes, see Table 2). Conversely, we did not observe any notable geographic patterns in model residuals (Fig. S6).

@@

The most important covariates in these models were those relating to spatial location, waterbody area, and perimeter (Fig. \ref{fig5}). Conversely, watershed metrics and waterbody elevation had little contribution to random forest model fit. The spatial location (i.e. HUC4, hydrologic subbasin) covariate was notably less important in the true slope model compared to the two proxy slope models. To evaluate the contribution of our “offset” models relative to the “base” geometric model, we can look at model importance calculations for the geometric max depth input to the random forest model (Fig. \ref{fig5}). These calculations indicate that omitting a geometric max depth term results in a 130%, 60%, or 50% increase in mean square error depending on the formulation of geometric max depth in Eq. \ref{eq1}.

@@

We found other differences among waterbody geometry measures according to waterbody cross-section shape. One finding was that convex waterbodies, when compared to concave waterbodies, had longer distances to waterbody centers relative to corresponding distances to the deepest point of waterbodies. In addition, convex waterbodies often had steeper in-lake slopes relative to nearshore land slopes as compared to concave waterbodies. Finally, it was notable that convex waterbodies were deeper than concave waterbodies despite having similar distributions of waterbody surface area (Fig. S8). The underlying cause of these differences is unknown but one possibility is that geometry is tied to the circumstances of waterbody formation whereby the formation of concave waterbodies were a result of more intense glacial scouring compared to that of convex waterbodies \citep{gorhamPhysicalLimnologyNorthern1958}. While our findings provide some evidence in support of this idea, namely that there is a geographic hotspot of concave waterbodies associated with the glaciated “prairie pothole region” (see \citealt{hayashiSimpleEquationsRepresent2000}), the overall geographic distribution of waterbody cross-section shapes does not support this idea. Instead of a concentrated area of concave waterbodies in formerly glaciated regions, there appears to be a fairly even mix of concave and convex waterbodies distributed amongst the northern (i.e. glaciated) and southern (non-glaciated) portions of our study area (Fig. \ref{fig2}).

### 2021-11-10 manuscript/si_template_2019.tex
stash progress

\authors{J. Stachelek\affil{1,2}, P. J. Hanly\affil{1}, and P. A. Soranno\affil{1}}

\affiliation{1}{Department of Fisheries and Wildlife, Michigan State University, 480
      Wilson Rd., East Lansing, MI 48824, USA}
\affiliation{2}{Center for Limnology, University of Wisconsin – Madison, Madison, WI, USA}

>>>>

\authors{J. Stachelek\affil{1,2,3}, P. J. Hanly\affil{3}, and P. A. Soranno\affil{3}}

\affiliation{1}{Earth and Environmental Sciences Division, Los Alamos National Laboratory, Los Alamos, NM, USA}
\affiliation{2}{Center for Limnology, University of Wisconsin – Madison, Madison, WI, USA}
\affiliation{3}{Department of Fisheries and Wildlife, Michigan State University, 480
      Wilson Rd., East Lansing, MI 48824, USA}
